Multidimensional Persistence Module Classification via Lattice-Theoretic Convolutions

### 2020-11-28 paper/paper.tex
update information about simulations, new plot

\paragraph{Remark}
Traditional convolutional neural networks are useful in part because the
convolution kernels (here the functions $g$) can have very small support,
reducing the number of parameters that must be learned. In the standard
convolutional setting, these kernels are implicitly supported in a neighborhood
of the origin, but the location of the kernel is not usually explicitly
specified. In the lattice setting, we do need to specify where the kernel
resides. In the abelian group case, the kernel is supported near the identity,
and similarly, when we treat our domain as a lattice, the kernel should be
supported near the neutral element of the operation. That is, for a meet
convolution, $g$ should be supported at the maximum $(m,n)$, and for a join
convolution, $g$ should be supported at the minimum $(0,0)$. This ensures that
the convolution operators are capable of preserving information at every point
of the space. For instance, if $g(0,0) \neq 0$, then $(f \ast_{\vee} g)(x,y)$ is
a sum of terms including $f(x,y)g(0,0)$, so all information from one layer can
be passed to the next layer. 
\section{Experiments}
We use a large portion of the Princeton ModelNet
dataset~\cite{zhirong_wu_3d_2015} as a source of finite
metric spaces. This dataset consists of hundreds of 3-dimensional CAD models
representing objects from 10 classes. We select two of the classes and sample
points from the 3D models to produce finite metric spaces embedded in $\R^3$.
We then compute the corresponding multidimensional persistence modules, from
which we produce features used as an input to a convolutional neural net
classifier.

@@

We compare the performance of two convolutional networks on this classification
task. One uses the lattice-convolution based layers described in
Section~\ref{sec:latticeCNN}, and the other uses standard convolutional layers.
Each has three convolutional layers followed by two fully connected layers.
The lattice-based convolution layers are of the form $\alpha \cdot \text{MeetConv}(x) +
(1-\alpha)\cdot\text{JoinConv}(x)$ for a hyperparameter $\alpha \in [0,1]$. All
convolution kernels have dimension $4 \times 4$, hidden convolution layers have
16 features, and the final convolution layer has 8 features. The first two convolution
layers are followed by max-pooling layers with a $2 \times 2$ kernel.
The inner fully connected layer has 32 features. We use a cross entropy loss
function with a softmax in the final layer. The lattice-based convolutional architecture
is summarized in Figure~\ref{fig:architecture}.
The networks are trained using the
Adam gradient algorithm with learning rate $10^{-4}$ for 300 epochs. We reserve
10\% of the data for testing. Results are shown in Figure~\ref{fig:comparison}.

@@

@@

\begin{figure}[h!]
  \caption{Comparison learning curves for the lattice neural network and
    standard convolutional neural network.}\label{fig:comparison}
    \begin{center}
    %\includegraphics[width=0.75\textwidth]{training.png}
    \includegraphics[width=0.9\textwidth]{testing.png}
    \end{center}
\end{figure}

\section*{Acknowledgments}
The authors would like to thank Chris Wendler for replicating our original experiment. We would like to thank the reviewers for providing helpful feedback. The author [HR] is partially supported by Office of Naval Research (Grant No. N00014-1442-16-1-2010).
\small
\bibliographystyle{plain}
\bibliography{latticenn}

>>>>

\paragraph{Remark}
Traditional convolutional neural networks are useful in part because the
convolution kernels (here the functions $g$) can have very small support,
reducing the number of parameters that must be learned. In the standard
convolutional setting, these kernels are implicitly supported in a neighborhood
of the origin, but the location of the kernel is not usually explicitly
specified. In the lattice setting, we do need to specify where the kernel
resides. In the abelian group case, the kernel is supported near the identity,
and similarly, when we treat our domain as a lattice, the kernel should be
supported near the neutral element of the operation. That is, for a meet
convolution, $g$ should be supported at the maximum $(m,n)$, and for a join
convolution, $g$ should be supported at the minimum $(0,0)$. This ensures that
the convolution operators are capable of preserving information at every point
of the space. For instance, if $g(0,0) \neq 0$, then $(f \ast_{\vee} g)(x,y)$ is
a sum of terms including $f(x,y)g(0,0)$, so all information from one layer can
be passed to the next layer. A bit more is necessary to avoid degenerate
convolutions. A kernel supported in a geometrically small neighborhood of the
neutral element results in trivial receptive fields for most neurons. For
example, if $(x,y)$ is greater than the maximal element of the support of $g$,
$(f \ast_{\vee} g)(x,y)$ is simply a scalar multiple of $f(x,y)$. As a result,
we hypothesize that the appropriate support sets for lattice-convolutional kernels
are evenly-spaced sublattices including both $(0,0)$ and $(m,n)$.

\section{Experiments}
We use a portion of the Princeton ModelNet
dataset~\cite{zhirong_wu_3d_2015} as a source of finite
metric spaces. This dataset consists of hundreds of 3-dimensional CAD models
representing objects from 10 classes. We sample
points from the 3D models to produce finite metric spaces embedded in $\R^3$.
We then compute the corresponding multidimensional persistence modules, from
which we produce features used as an input to a convolutional neural net
classifier.

@@

We compare the performance of two convolutional networks on this classification
task. One uses the lattice-convolution based layers described in
Section~\ref{sec:latticeCNN}, and the other uses standard convolutional layers.
Each has three convolutional layers followed by two fully connected layers.
The lattice-based convolution layers are of the form $\alpha \cdot \text{MeetConv}(x) +
(1-\alpha)\cdot\text{JoinConv}(x)$ for a hyperparameter $\alpha \in [0,1]$. We
set $\alpha = \frac{1}{2}$. All
convolution kernels have dimension $4 \times 4$, hidden convolution layers have
16 features, and the final convolution layer has 8 features. The first two convolution
layers are followed by max-pooling layers with a $2 \times 2$ kernel. For the
lattice convolutional layers, the support of the kernel lay in an evenly spaced
$4 \times 4$ grid of points in $[m]\times [n]$, while the standard convolutional
layers had a traditional contiguously supported kernel.
The inner fully connected layer has 32 features. We use a cross entropy loss
function with a softmax in the final layer. The lattice-based convolutional architecture
is summarized in Figure~\ref{fig:architecture}.
The networks are trained using the
Adam gradient algorithm with learning rate $2 \times 10^{-4}$ for 300 epochs. We reserve
10\% of the data for testing. Results are shown in Figure~\ref{fig:comparison}.

@@

The lattice convolutional network slightly underperforms the standard
convolutional classifier. This is somewhat disappointing, and perhaps indicates
that the relevant information encoded in degree-0 persistent homology is best
captured with a geometric notion of locality in $\R^2$.\footnote{Due to a coding
  error, an earlier version of this paper gave results incorrectly showing the
  lattice convolutional network outperforming the traditional network.} This
work has not exhausted the possibilities for lattice convolutions in
multidimensional persistence, but does suggest circumspection in evaluating
their future use.

@@

\begin{figure}[h!]
  \caption{Comparison learning curves for the lattice neural network and
    standard convolutional neural network.}\label{fig:comparison}
    \begin{center}
    %\includegraphics[width=0.75\textwidth]{training.png}
      \includegraphics[width=\textwidth]{test_results_2.png}
    \end{center}
\end{figure}

\section*{Acknowledgments}
The authors would like to thank Chris Wendler for replicating our original
experiment. We would like to thank the reviewers for providing helpful feedback.
The author [HR] is partially supported by Office of Naval Research (Grant No.
N00014-1442-16-1-2010), and the author [JH] is supported by the National Science
Foundation (NSF-DMS \#1547357).
\small
\bibliographystyle{plain}
\bibliography{latticenn}
